---
title: "Experimental harvest of ptarmigan, black grouse and capercaillie:" 
subtitle: "Summary of planning stage"
first_author:
  - name: Erlend B. Nilsen
    address: NINA Trondheim
    email: erlend.nilsen@nina.no
second_authors:
  - name: Lasse F. Eriksen
    address: NINA Trondheim
    email: lasse.eriksen@nina.no
  - name: Christoffer Høyvik Hilde
    address: NINA Trondheim
    email: christoffer.hilde@nina.no
last_author:    
  - name: Matthew Grainger
    address: NINA Trondheim
    email: matthew.grainger@nina.no
reportnr: 2570
location: Trondheim
ISBN_end: 1234-5
self-ref: "Nilsen, E.B., Eriksen, L.F., Høyvik Hilde, C., Grainger, M.J. 2025. Experimental harvest of ptarmigan, black grouse and capercaillie: Summary of planning stage. NINA Rapport 2570"
english-self-ref: "Nilsen, E.B., Eriksen, L.F., Høyvik Hilde, C., Grainger, M.J. 2025. Experimental harvest of ptarmigan, black grouse and capercaillie: Summary of planning stage. NINA Report 2570"
handle-endpoint: [fås av bibliotek]
access: Åpen
secure_pdf:
cover_picture: Gordon.jpg
cover_picture_text: Forsidebildetekst
cover_picture_photographer: Erlend B. Nilsen
line_numbering:
qualitycontrol: [xx]
boss_sign: Forskningssjef [fylles ut av forskningssjefen] (sign.)
customer: [xx]
customer_ref: [xx]
customer_contact: [xx]
bibliography: mybibfile.bib
csl: nina.csl
format:
  # ninareport-pdf:
  #   keep-tex: true
  #   highlight: tango
  #   fig_caption: yes
  #   latex_engine: xelatex
  #  # default-image-extension: svg
  #   fig-format: pdf
  #   fig-dpi: 600
  ninareport-html:
    highlight-style: tango
    toc: true
    toc-depth: 3
    number-sections: true
    toc-location: left
    #default-image-extension: svg
    fig-format: svg
    toc-title: Innhold
  # ninareport-docx:
  #   highlight: tango
  #   #prefer-html: true
  #   toc: true
  #   toc-depth: 3
  #   number-sections: true
crossref:
  fig-title: Figur     # (default is "Figure")
  tbl-title: Tabell     # (default is "Table") 
  fig-prefix: Figur   # (default is "Figure")
  tbl-prefix: Tabell    # (default is "Table")   
  sec-prefix: Kapittel
foreword: 
    text: [Forord inn her]
    location: [Sted, forord]
    name: [Prosjektleder]
sammendrag: |
  Tekst inn her, et kort resymé av innholdet. \textbf{Maksimalt 4000 tegn inkl. ordmellomrom. Dette er maxgrense i sammendragsfeltet i CRIStin, og dersom sammendraget er lengre kommer det ikke med her.} Teksten i sammendraget er søkbar i databaser og på nett, og er viktig for at rapporten skal fanges opp ved søk.\
  \
  Seksjoner og tomme rader mellom dem er litt tricky å få til i YAML-avsnittet, men det kan gjøres slik. Tomme rader i forordet kan også lages på samme måte.
  
abstract: Tekst inn her, et kort resymé av innholdet. \textbf{Maksimalt 4000 tegn inkl. ordmellomrom. Dette er maxgrense i sammendragsfeltet i CRIStin, og dersom sammendraget er lengre kommer det ikke med her.} Teksten i sammendraget er søkbar i databaser og på nett, og er viktig for at rapporten skal fanges opp ved søk.
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

require(NinaR)
require(xtable)
require(knitr)
require(tidyverse)
require(ggpubr)
library(rgbif)
library(purrr)


```

# Introduction {#inl}

Sustainable harvest management of our game resources is a key goal in Norwegian nature management. In order to achieve sustainable management, it is generally recommended to an adaptive approach, where one actively searches for new knowledge that is necessary to fill existing knowledge gaps [@Johnson1997]. In recent years, however, people have become more aware of the partly enormous changes we face in the coming years, which are expected to largely affect hunting management. These challenges are largely linked to climate change and other (man-made) changes in the natural environment, but also changes in other components of the socio-ecological systems wildlife species are part of are expected to influence how sustainable management is defined [@Bowler2020; @Eriksen2018; @Henden2020; @Mansson2023]. With increased focus in society on the loss of biological diversity on a global and national scale, one can further expect increased expectations in society that recreational hunting is carried out in a sustainable way with little risk of overharvesting. This will lead to stricter requirements for documentation of sustainability and data bases will increase. More recent conceptual studies of adaptive management have discussed how such new "settings" not only force new knowledge needs, but that it can also lead to changed objectives and measurement parameters for adaptive management [@Mansson2023].

A central concept for predicting the effect of hunting on wildlife populations is the relationship between additive hunting mortality (which comes in addition to other mortality) vs compensatory hunting mortality (which is compensated by other mortality being reduced in proportion to the harvest mortality - see @Perone2013 and @Sandercock2011). Our knowledge of this comes largely from experimental studies carried out at the end of the 1990s under the auspices of the "Rypejaktprosjektet" (with Hans Chr. Pedersen, NINA, as project leader). Here it was concluded that hunting mortality was partially compensatory at low harvest rates, while it was additive or even super-additive at high harvest rates [@Sandercock2011; @Pedersen2004]. This knowledge base was produced by carrying out large-scale experimental studies, where the hunting harvest (respectively 0, 15 and 30%) was distributed randomly between the study areas. Effects were investigated both directly in terms of survival (using radio-tagged birds - see @Sandercock2011) and indirectly via population densities (see @Pedersen2004). This project focused exclusively on willow ptarmigan (*Lagopus lagopus*), and no similar knowledge exist for black grouse *(Lyrurus tetrix*) and capercaillie (*Tetrao urogallus*).

Whether results from the former large scale experiement is transferable in time (i.e. to the current environmental conditions) and to other species (i.e. black grouse and capercaillie) is an empirical question that cannot be answered by directly by existing data. For example, one can expect that climate change will have significant effects on the trophic interactions that largely affect grouse population dynamics and densities [@Bowler2020, @Henden2020]. Changes in these interactions will have the potential to directly affect grouse populations' potential to compensate for hunting mortality (see discussions in e.g. @Perone2013). Similarly, large changes in the size of the grouse population could have similar effects, since the degree of compensation can be closely linked to density dependence [@Sinclair1996]. One way to address these uncertainties would be to conduct an conceptual replication of the well-known experiments carried out in the 90-ties. Such conceptual replication studies (*sensu* @Nakagawa2015) are rare in ecology [@Fraser2020], but have recently been strongly recommended to examine the validity and generalizability of previously conducted studies [@Fraser2020; @Nakagawa2015].

However, before such a replication is carried out it is a prerequisite that an up-to-date mapping of the research-based literature on the sustainable management of galliformes from a climate change perspective is conducted (see also @Grainger2020), to ensure that the replication rests on the best available knowledge. In addition, a thorough planning of how best to design such in a way that both sample sizes and climatic gradients are sufficient to close the most pressing knowledge gaps. This report is a part of the planning stage of such a project, where the intent is to apply experimental harvest treatment to populations of willow grouse, black grouse and capercaillie in Norway. It includes two main sections:

-   A systematic mapping of the literature related to effects of harvest on galliformes

-   A thorough evaluation of the experimental design including both the sampling of study subjects and assignment of treatment levels.

We will first present the results from the systematic mapping of the literature, and then the evaluation of experimental design. In the final chapter, we will present a detailed description of the suggested design for the new harvest experiment.

# Systematic mapping of the literature

This systematic map is produced as part of the preparation of a large-scale harvest experiment in Norway outlined above. We aimed to identify studies that have quantitatively assessed the effects of legal harvest (i.e. managed hunting) on Galliformes to ensure that the planned experimental studies on the sustainability of galliform hunting are based on the best available evidence. This mapping aims to identify, categorise, and synthesise existing research on how hunting mortality interacts with ecological and environmental factors. By systematically compiling relevant studies, we provide a foundation for assessing knowledge gaps and determining the generalisability of past research.

The [protocol](https://osf.io/ny2jt/) was prepared prior to conducting the search. Here, we include the relevant information and report any deviation from the protocol.

## Objective of the review and research question

The main objective of this review is to map the available literature on effects of managed harvest/hunting on species of the order Galliformes.

## Methods 

### Search for articles 

We used a PIO (Population, Intervention, Outcome) question structure to help develop the search strategy.  

-   Population (P): Bird species within the order Galliformes 

-   Intervention (I): Legal (managed) hunting or harvest 

-   Outcome (O): The ecological impacts of (managed) hunting on population dynamics, demography, genetics or behaviours  

We searched for academic literature (publications) in two search platforms; Web of Science and lens.org. Specifically, we searched the Web of Science Core Collection (WOS.SCI: 1987 to 2025, WOS.AHCI: 1987 to 2025, WOS.ESCI: 2020 to 2025 and WOS.SSCI: 1987 to 2025) using the NINA library subscription and the following search string: 

((TS=(galliform\* OR grouse OR ptarmigan OR Phasianidae OR pheasant OR "lagopus lagopus" OR "lagopus muta" OR "wild turkey" OR "bobwhite" OR partridge)) 

AND (TS=("hunt" OR "harvest"))) 

AND SU=("Ecology" OR "Wildlife" OR "Environment") 

For lens.org we used the following search string: 

(abstract:("galliform" OR "grouse" OR "ptarmigan" OR "Phasianidae" OR "pheasant" OR "lagopus lagopus" OR "lagopus mura" OR "wild turkey" OR "bobwhite" OR "partridge") OR title:("galliform" OR "grouse" OR "ptarmigan" OR "Phasianidae" OR "pheasant" OR "lagopus lagopus" OR "lagopus muta" OR "wild turkey" OR "bobwhite" OR "partridge")) AND abstract:("hunt" OR "harvest") OR title:("hunt" OR "harvest") AND field_of_study:("Ecology" OR "Wildlife" OR "Environment") 

In addition, a backward and forward citation chasing (using CitationChaser) was performed based on the three first benchmark papers listed below [@Sandercock2011; @Pedersen2004; @Sedinger2005].  

### Benchmark papers 

The following benchmark papers was used to assess the effectiveness of the search string. We expected to find all these papers in a review of this question. We recorded if each benchmark paper is found in the searches (WoS and lens.org).  

-   Sandercock, B. K., E. B. Nilsen, H. Brøseth, and H. C. Pedersen. 2011. Is hunting mortality additive or compensatory to natural mortality? Effects of experimental harvest on the survival and cause-specific mortality of willow ptarmigan. Journal of Animal Ecology 80:244-258. 

-   Pedersen, H. C., H. Steen, L. Kastdalen, H. Brøseth, R. A. Ims, W. Svendsen, and N. G. Yoccoz. 2004. Weak compensation of harvest despite strong density-dependent growth in willow ptarmigan. Proceedings of the Royal Society of London Series B-Biological Sciences 271:381-385. 

-   Sedinger, J.S. & Rotella, J.J. 2005. Effect of harvest on sage-grouse Centrocercus urophasianus populations: what can we learn from the current data? Wildlife Biology, 11, 371–375. 

-   Pedersen, Å. Ø., E. M. Soininen, S. Unander, M. H. Willebrand, and E. Fuglei. 2014. Experimental harvest reveals the importance of territoriality in limiting the breeding population of Svalbard rock ptarmigan. European Journal of Wildlife Research 60:201-212. 

-   Brøseth, H., J. Tufto, H. C. Pedersen, H. Steen, and L. Kastdalen. 2005. Dispersal patterns in a harvested willow ptarmigan population. Journal of Applied Ecology 42:453-459. 

-   Brøseth, H., E. B. Nilsen, and H. C. Pedersen. 2012. Temporal quota corrections based on timing of harvest in a small game species. European Journal of Wildlife Research 58:797-802. 

-   Brøseth, H., and H. C. Pedersen. 2000. Hunting effort and game vulnerability studies on a small scale: a new technique combining radio-telemetry, GPS and GIS. Journal of Applied Ecology 37:182-190. 

-   Henden, J.-A., R. A. Ims, N. G. Yoccoz, E. J. Asbjørnsen, A. Stien, J. P. Mellard, T. Tveraa, F. Marolla, and J. U. Jepsen. 2020. End-user involvement to improve predictions and management of populations with complex dynamics and multiple drivers. Ecological Applications n/a:e02120. 

-   Small, R.J., Holzwart, J.C. & Rusch, D.H. 1991. Predation and hunting mortality of ruffed grouse in central Wisconsin. Journal of Wildlife Management, 55, 512–520. 

In our searches, all benchmark papers were found except.... XXX, which was not indexed by any of the databases that we searched. Overall, this is found to be satisfactory.

### Article screening  

We used pre-determined inclusion/exclusion criteria when screening eligible studies. Excluded papers will be shown in a flowchart using the ROSES flow chart [@Haddaway2018]. Included studies was de-duplicated based on DOI and title matches using the deduplication tool in Rayyan.  

After de-duplication, title and abstract screening was done using the eligibility criteria. If uncertainty remained, the full text version was visited to confirm/reject. The screening was done using Rayyan. A set of 54 papers was reviewed by three members of the review team to ensure consistency.  

### Eligibility criteria 

We reviewed scientific articles from all parts of the world. We included papers covering all galliform species. Studies of both native and introduced populations were eligible. Studies of all spatial scales and across all relevant ecosystems were considered eligible. Only studies documenting the effects of legal and/or managed harvest on galliform populations were eligible. Studies of poaching were not eligible. The effects must be assessed quantitatively for the study to be eligible. Eligible outcomes included effects of harvest on population abundance, density, sex- and age structure, recruitment, survival, behaviour, genetics, physiology and distribution. Studies focusing on lead (or heavy metal) poisoning or content in wild birds from shooting were not eligible. Both observational and experimental study designs are eligible. Mathematical simulation studies are eligible.    

### Title and abstract screening stage 

We used a decision tree to make decisions about inclusion and exclusion at the Title and Abstract stage. The tree is presented in the protocol and in the Appendix to this report. The decision tree helped the reviewers to make decisions about eligibility of each paper in a structured way. Duel screening was carried out on 20 % of papers (n = 183). Agreement between reviewers was assessed using the kappa statistic which was over 0.7.

### Full text eligibility 

Papers included at the Title and Abstract stage was taken forward to the full text screening stage. A RIS file of included articles was downloaded from Rayyan and added to a Zotero database. PDF versions of each paper was searched using Zotero’s “Find Available PDF” tool, ResearchGate, Google Scholar and Google. If a PDF was not available online (or was not open to us), the paper was excluded at this stage of the review. Full text papers was excluded if they are/did not: 

-   Written in English, Norwegian, Swedish or Danish  
-   Focus on species from order Galliformes
-   Document the effect of legal or managed hunting on Galliformes
-   Assess the effect of hunting quantitatively 

### Data coding  

From all papers that met the eligibility criteria, we extracted key variables (aka “data coding”). In addition to metadata about the publication (publication year, journal name etc) we extracted from the full-text information about study species, study area location and extent, research approach and experimental design, outcome variables and to which extent the paper reported harvest rates and to which extent the study assessed whether harvest was additive or compensatory. For outcome variables, we used the Essential Biodiversity Variable (EBV) framework (refs), and coded both for EBV class (high order classification) and EBV name (lower order classification). A detailed description of data coding is presented in the protocol and in the Appendix.    

## Results from the literature review

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

library(googlesheets4)
# Read data
#data <- read_sheet("https://docs.google.com/spreadsheets/d/1Shdb7vfLn-bnZd0-FT2Ngqovxi1iEn-MYLnTtBPpBQ4/edit?resourcekey=&gid=968606953#gid=968606953")
#

#saveRDS(data,"data/GS_data_Raw.RDS")
data <- readRDS("data/GS_data_Raw.RDS")



data<-data |> 
  distinct(`Study ID`, .keep_all = TRUE)

excluded <- data |>
  filter(!rowSums(across(4:8, ~ . == "Yes")) == 5)
included <- anti_join(data, excluded)
```

Our searches returned 1023 articles in total, with 542 from Web of Science, 112 from lens.org, 9 benchmark articles and 192 from backward-, and 168 from forward citation chasing from the three benchmark articles mentioned above [@Sandercock2011; @Pedersen2004; @Sedinger2005]. Of these 173 were duplicates which were removed from the database. We screened the remaining 850 articles for eligibility at the Abstract and Title stage. Of the 850 included articles, a total of 200 were moved further to full-text screening, of which 166 papers that had an "include" decision and 34 a "maybe" decision. There were 9 conflicts (where reviewers disagreed on eligibility). All these papers were taken through to the fulltext stage and assessed for eligibility. Exclusion decisions (650 records) were due to the research not focusing on the effects of legal or managed hunting (365 records), not being focused on Galliformes (276 records), not quantitative (23 records), abstract not found online (16 records), or abstract not available in English, Norwegian, Swedish or Danish (3 records).

At the fulltext stage `r dim(excluded)[1]` articles were excluded (3 with no available pdf, 6 not focused on Galliformes, 12 not assessing the effects of hunting quantitatively and the rest, 61, not documenting the effects of legal harvest). This left `r dim(included)[1]` papers that went forward to the data extraction stage.

![ROSES Flowchart showing the number of papers included or excluded in the review](images/roses%20(3)-01.png){#fig-prisma}

```{r}
#| echo: false
#| warning: false
#| message: false

library(synthesisr)
all_RIS <- read_refs("data/articles.ris")

included_full_details <- left_join(included, all_RIS, by = c("Study ID" = "accession_zr"))

included_full_details <- included_full_details |> janitor::clean_names()
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: colour palette 
species_colours <- c(
  "Centrocercus urophasianus" = "#736F4C",  # Sage-grouse (greyish-brown)
  "Dendragapus obscurus" = "#5A4A42",       # Dusky Grouse (brown-grey)
  "Lagopus lagopus" = "#85361C",            # Willow Ptarmigan (reddish-brown)
  "Lyrurus tetrix" = "#1C1C1C",             # Black Grouse (black)
  "Tetrao urogallus" = "#2B2B2B",           # Capercaillie (dark grey/black)
  "Tetrastes bonasia" = "#8C7051",          # Hazel Grouse (brownish-grey)
  "Perdix perdix" = "#A67B5B",              # Grey Partridge (grey-brown)
 "Lagopus muta" = "#6E6259",               # Rock Ptarmigan (darker grey-brown)
  "Alectoris graeca" = "#B0A48B",           # Rock Partridge (grey-brown)
  "Alectoris rufa" = "#A5694F",             # Red-legged Partridge (red-brown)
  "Alectoris chukar" = "#B48B7D",           # Chukar Partridge (beige-brown)
  "Bonasa umbellus" = "#7A5F47",            # Ruffed Grouse (brown-grey)
  "Colinus virginianus" = "#9E8358",        # Northern Bobwhite (tan-brown)
  "Meleagris gallopavo" = "#4A2D1D",        # Wild Turkey (dark brown/bronze)
  "Phasianus colchicus" = "#B22222"         # Common Pheasant (red and green)
)

```

Legal hunting of Galliformes has been the subject to extensive research. In our searches we identified studies published as far back as 1963, but we are aware that research on this topic was carried out even before that, however these studies were not covered by our searches. The number of studies per year peaked in 2020 (8 studies per year) and the long-term average was 2.93 studies per year (@fig-Publication_year).

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-Publication_year
#| fig-cap: "The number of studies per year"

included_full_details |>
  group_by(year) |>
  tally() |>
  drop_na() |>
  ggplot(aes(year, n)) +
  geom_histogram(stat = "identity", fill=species_colours[3]) + # colour of a Willow Ptarmigan
  labs(
    x = "Year",
    y = "Number of records",
    #fill = "Response"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        text = element_text(size =13))
```

The most common geographic location for studies was in the USA (99 studies), followed by Norway (16) and the UK (9) (@fig-location). In the USA studies were carried out in 33 States (Alaska, California, Colorado, Florida, Georgia, Idaho, Illinois, Indiana, Kansas, Kentucky, Louisiana, Maine, Michigan, Mississippi, Missouri, Montana, Nevada, New York, North Carolina, Ohio, Oklahoma, Oregon, Pennsylvania, Poland, Rhode Island, South Carolina, South Dakota, Tennessee, Texas,Utah,Virginia, West Virginia, Wisconsin).

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-location
#| fig-cap: "Geographic location of included studies"

included_full_details |>
  separate_rows(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma, sep = ", ") |> # Split at comma and create new rows
  group_by(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma) |>
  mutate(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma = case_when(
    study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma %in% c(
      "Alaska", "California", "Colorado", "Florida", "Georgia","Georiga", "Georgia (USA)", "Idaho", "Illinois", "Indiana", "Kansas", "Kentucky", "Louisiana", "Maine", "Michigan", "Mississippi", "Missouri", "Montana", "Nevada", "New York", "North Carolina", "Ohio",
      "Oklahoma", "Oregon", "Pennsylvania", "Poland", "Rhode Island", "Rhode island", "South Caolina", "South Caraolina", "South Carolina", "South Dakota", "Tennessee", "Texas", "Texas,Texas", "US", "Utah",
      "Virgina", "Virginia", "West Virginia", "Wisconsin", "Wisonsin"
    ) ~ "USA",
    TRUE ~ study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma
  )) |> 
  mutate(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma= case_when(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma %in% c("United Kingdom", "UK", "Scotland", "England")~"UK",
            TRUE~study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma)) |> 
tally() |>
  drop_na() |>
  ggplot(aes(reorder(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma, n), n)) +
  geom_histogram(stat = "identity", fill=species_colours[3]) +
  labs(
    x = "Study area location",
    y = "Number of records"
  ) +
  #ggthemes::theme_base() +
  theme_minimal()+
  coord_flip() +
  theme(text = element_text(size = 13))
```

The extent of the study areas were predominately regional or local, with very few studies at the site level (@fig-Study_area). Few studies were of "national" or multinational scale. Note that following our definition used here, regional studies typically cover one mountain area or a county, whereas a local study would cover a municipality or parts of a municipality. Note that this contrasts previous calls for more large-scale research [@estes_spatial_2018] that might be need to uncover the ecological dynamics and provide input to management and natural resource management policy at larger spatial scales.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-Study_area
#| fig-cap: "The study area extent"

included_full_details |>
  group_by(study_extent_size_of_the_study_area) |>
  tally() |>
  drop_na() |>
  #ggplot(aes(reorder(study_extent_size_of_the_study_area, n), n)) +
  mutate(study_extent_size_of_the_study_area = fct_recode(study_extent_size_of_the_study_area,
             Regional = "Regional (e.g. a county, a larger mountain area etc)", 
             Local = "Local (e.g. a municipality or part of municipality)", 
             `Site level` = "Site level (< 1km2)")) |>
  mutate(name = fct_relevel(study_extent_size_of_the_study_area, 
                            "Multinational", "National", "Regional", "Local", "Site level")) |>
  ggplot(aes(x=name, y=n)) +
  geom_histogram(stat = "identity", fill="#85361C") +
  labs(
    x = "Spatial extent",
    y = "Number of records"
  ) +
  theme_minimal() +
  coord_flip() +
  theme(text = element_text(size = 15))

```

Most studies covered in our mapping were observational (106 studies), but there were also a substantial number of studies that relied on simulations or scenario analysis (35 records; @fig-Research_design a)). Note that the sum of observational studies based on empirical data and studies based on simulations is higher than the total of 129 papers that were found eligible. This is because some studies used both research approaches and thus were included in both categories. In agreement with previous assessments of study designs in ecological research (refs) most of the studies that presented empirical data were based on simple (and less robust) study designs (@fig-Research_design b)). There were only 6 papers using a "Randomized Controlled Trial (RCT)" approach, the study design that leads to the highest level of evidence for causal understanding [@christie2019simple] and these were limited to Norway (5) and the USA (1). The Norwegian articles are all from the same research project/experiment conducted in the late 1990-ties.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-Research_design
#| fig-cap: "Type of research carried out"

p1 <- included_full_details |>
  separate_rows(research_type, sep = ", ") |> # Split at comma and create new rows
  group_by(research_type) |>
  tally() |>
  drop_na() |>
  mutate(research_type = fct_recode(research_type, `Empirical data` = "Observational / analysis of empirical data", `Simulation based` = "Simulation based /Scenarios")) |>
  ggplot(aes(reorder(research_type, n), n)) +
  geom_histogram(stat = "identity", fill=species_colours[2]) +
  labs(
    x = "Research type",
    y = ""
  ) +
  theme_minimal() +
  coord_flip() +
  theme(text = element_text(size = 15))


p2 <- included_full_details |>
  group_by(study_design_observational_studies_skip_if_simulation_study_selected_above) |>
  tally() |>
  drop_na() |>
  mutate(study_design_observational_studies_skip_if_simulation_study_selected_above = fct_recode(study_design_observational_studies_skip_if_simulation_study_selected_above, BACI = "Before-After-Control-Impact (BACI)", RCT = "Randomized Controlled Trials (RCTs)")) |>
mutate(name = fct_relevel(study_design_observational_studies_skip_if_simulation_study_selected_above, 
                            "RCT", "BACI", "Before-After (BA)", 
                            "Control-Impact (CA)", "After (A)")) |>
  ggplot(aes(x=name, y=n)) +
  geom_histogram(stat = "identity", fill=species_colours[3]) +
  labs(
    x = "Study design",
    y = "Number of Responses"
  ) +
  theme_minimal() +
  coord_flip() +
  theme(text = element_text(size = 15))

ggarrange(p1, p2, ncol = 1, labels = c("a)", "b)"), 
          align= "v")



```

```{r}
#| echo: false
#| eval: false
#| warning: false
#| message: false
#| label: fig-Study_design_by_country
#| fig-cap: "Study design by country"

country_data<-included_full_details |>
  separate_rows(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma, sep = ", ") |> # Split at comma and create new rows
  group_by(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma) |>
  mutate(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma = case_when(
    study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma %in% c(
      "Alaska", "California", "Colorado", "Florida", "Georgia","Georiga", "Georgia (USA)", "Idaho", "Illinois", "Indiana", "Kansas", "Kentucky", "Louisiana", "Maine", "Michigan", "Mississippi", "Missouri", "Montana", "Nevada", "New York", "North Carolina", "Ohio",
      "Oklahoma", "Oregon", "Pennsylvania", "Poland", "Rhode Island", "Rhode island", "South Caolina", "South Caraolina", "South Carolina", "South Dakota", "Tennessee", "Texas", "Texas,Texas", "US", "Utah",
      "Virgina", "Virginia", "West Virginia", "Wisconsin", "Wisonsin"
    ) ~ "USA",
    TRUE ~ study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma
  ))

# Aggregate data to get counts
heatmap_data <- country_data %>%
  count(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma, study_design_observational_studies_skip_if_simulation_study_selected_above)|> 
  drop_na()

# Rename columns for easier use in ggplot
colnames(heatmap_data) <- c("Country", "Study_design", "Count")
ggplot(heatmap_data, aes(x = Country, y = Study_design, fill = Study_design)) +
  geom_tile(aes(alpha=Count)) +
 #scale_fill_manual(values = species_colours) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for readability
  labs(x = "Country",
       y = "Study design",
       fill = "")
```

We also extracted information about to which extent harvest rates (or percentage of the population removed by harvest) were reported in 71 papers. The remaining papers did not explicitly report such information (@fig-Harvest_rateAdditive a)). Moreover, among the papers that reported harvest rate or harvest mortality, only 27 made any assessment to which extent harvest was additive or not (@fig-Harvest_rateAdditive b)). Note however that this is somewhat ambiguous, and it might depend on how the concept of additivity is defined: although we did not quantify this by direct coding we note that much fewer studies were in fact able to properly document additivity vs compensation via a robust quantitative approach (see also discussion in ref).

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-Harvest_rateAdditive
#| fig-cap: "Reporting of harvest additivity"

p3 <- included_full_details |>
  group_by(is_harvest_rate_or_harvest_mortality_estimated_and_reported) |>
  tally() |>
  drop_na() |>
  ggplot(aes(x = is_harvest_rate_or_harvest_mortality_estimated_and_reported, y = n)) +
  geom_histogram(stat = "identity", fill=species_colours[3]) +
  labs(
    x = "Harvest rate estimated",
    y = "Number of records"
  ) +
  theme_minimal() +
  ylim(0, 100) + 
  theme(text = element_text(size = 15))



p4 <- included_full_details |>
  group_by(is_there_an_assessment_of_additivity_of_harvest_mortality) |>
  tally() |>
  drop_na() |>
  ggplot(aes(x = is_there_an_assessment_of_additivity_of_harvest_mortality, y = n)) +
  geom_histogram(stat = "identity", fill=species_colours[2]) +
  labs(
    x = "Assessment of additivity",
    y = ""
  ) +
  theme_minimal() +
  ylim(0, 100) + 
  theme(text = element_text(size = 15))

ggarrange(p3, p4, ncol = 2, labels = c("a)", "b)"), 
          align= "h")


```

From each of the eligible studies we assessed which outcome variables were examined (@fig-EBV_name2), at two different levels (EBV class and EBV names, respectively). In general, there were about equal number of studies focusing on "Species traits" (67 records) and "Species populations" (60 records). Note that we did not record this variable for simulation based studies. Separating the EBV classes into EBV names, we found that the most commonly studies outcome variable was "Survival" (67 records), followed by "Population density/abundance/growth\*" (47 records). In addition, several studies focused on effects of "Movement rates" or "Habitat use/selection" (12 and 6 records, respectively), and 11 papers reported effects of harvest on "Sex- and age-structure".

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-EBV_name2
#| fig-cap: "The number of studies for each EBV name"


included_full_details |>
  separate_rows(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework, sep = ", ") |>
  group_by(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework) |>
  mutate(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework = case_when(
    outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework %in% c("Behaviour", "Display behaviour/ Gobbling", "Foraging behaviour", "Gobbling activity") ~ "Behaviour",
    TRUE ~ outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework
  )) |>
  mutate(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework = case_when(
    outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework %in% c("Population density *", "Species abundance", "Population growth rate") ~ "Population density/abundance/growth*",
    TRUE ~ outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework
  )) |>
  tally() |>
  drop_na() |>
  mutate(`EBV class` = if_else(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework == "Population density/abundance/growth*", "Species population", "Species trait")) |>
  mutate(`EBV class` = if_else(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework == "Sex- and age structure*", "Species population", `EBV class`)) |>
  ggplot(aes(reorder(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework, n), n)) +
  geom_bar(stat = "identity", aes(fill = `EBV class`)) +
  #geom_histogram(stat = "identity", fill=species_colours[9]) +
  scale_fill_manual(values=c("#5A4A42", "#85361C")) +
  labs(
    x = "EBV name",
    y = "Number of records"
  ) +
  theme_minimal() +
  coord_flip() + 
  theme(legend.position = "top", legend.title = element_blank()) + 
  theme(text = element_text(size = 15))



```

Among studies that were based on (mathematical or statistical) simulations, the most common approach was to compare different harvest strategies (22 records), followed by conducting population viability analysis (17 records) and various forms of sensitivity analysis (13 records). In total, 11 studies were aiming to estimate/calculate maximum sustainable yield. Note that many studies used several of these approaches, so that the combined numbers will be higher than the total of 35 papers that included simulations or scenario analysis.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-simulations
#| fig-cap: "The purpose of simulation studies"

included_full_details |>
  separate_rows(for_simulation_based_scenario_studies_what_is_the_purpose_of_the_simulations, sep = ", ") |> # Split at comma and create new rows
  group_by(for_simulation_based_scenario_studies_what_is_the_purpose_of_the_simulations) |>
  tally() |>
  drop_na() |>
  ggplot(aes(reorder(for_simulation_based_scenario_studies_what_is_the_purpose_of_the_simulations, n), n)) +
  geom_histogram(stat = "identity", fill=species_colours[3]) +
  labs(
    x = "What is the purpose of simulations",
    y = "Number of records"
  ) +
  theme_minimal() +
  coord_flip() + 
  theme(text = element_text(size = 15))

```

The number of studies were very unequal among species (@fig-species): The most frequently studied species were wild turkeys (*Meleagris gallopavo*; 33 records), followed by willow ptarmigan/grouse (*Lagopus lagopus*; 28 records), bobwhite (Colinus virginianus; 18 records) and sage grouse (*Centrocercus urophasianus*; 18 records). The remaining 11 species that were represented among the eligible papers each had less than 10 papers. Both capercaillie (*Tetrao urogallus*; 4 records) and black grouse (*Lyrurus tetrix*; 6 records) are among these poorly studied species.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-species
#| fig-width: 15
#| fig-cap: "The number of studies for each species and country"


country_data<-included_full_details |>
  separate_rows(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma, sep = ", ") |> # Split at comma and create new rows
  group_by(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma) |>
  mutate(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma = case_when(
    study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma %in% c(
      "Alaska", "California", "Colorado", "Florida", "Georgia","Georiga", "Georgia (USA)", "Idaho", "Illinois", "Indiana", "Kansas", "Kentucky", "Louisiana", "Maine", "Michigan", "Mississippi", "Missouri", "Montana", "Nevada", "New York", "North Carolina", "Ohio",
      "Oklahoma", "Oregon", "Pennsylvania", "Poland", "Rhode Island", "Rhode island", "South Caolina", "South Caraolina", "South Carolina", "South Dakota", "Tennessee", "Texas", "Texas,Texas", "US", "Utah",
      "Virgina", "Virginia", "West Virginia", "Wisconsin", "Wisonsin"
    ) ~ "USA",
    TRUE ~ study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma
  ))|> 
  mutate(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma= case_when(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma %in% c("United Kingdom", "UK", "Scotland", "England")~"UK",
            TRUE~study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma))

country_sp_data<-country_data |> 
  separate_rows(species_if_multiple_seperate_with_a_comma_use_full_latin_names_e_g_lagopus_lagopus_lagopus_muta, sep = ", ") |>
  mutate(
    species_corrected = map_chr(
      species_if_multiple_seperate_with_a_comma_use_full_latin_names_e_g_lagopus_lagopus_lagopus_muta,
      ~ {
        res <- tryCatch(name_backbone(name = .x), error = function(e) NULL)
        if (!is.null(res) && "species" %in% names(res) && !is.na(res$species)) {
          res$species # Return corrected name
        } else {
          .x # Return original if not found
        }
      }
    )
  ) 

# Aggregate data to get counts pr species
species_data <- country_sp_data |> group_by(species_corrected) |>
  tally() |>
  drop_na() |>
  rename(Species = species_corrected)

# Aggregate data to get counts pr species pr country
heatmap_data <- country_sp_data %>%
  count(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma, species_corrected)|> 
  drop_na()

colnames(heatmap_data) <- c("Country", "Species", "Count") 
heatmap_data <- heatmap_data %>% left_join(., species_data)

p7 <- ggplot(species_data, aes(x = reorder(Species, n), y = n, fill = Species)) +
    geom_col() +
    scale_fill_manual(values = species_colours) + 
    labs(
        x = "Species",
        y = "Total number of records"
    ) +
    theme_minimal() +
    coord_flip() + 
    theme(text = element_text(size = 20))



p8 <- ggplot(heatmap_data, aes(x = Country, y = reorder(Species, n), size = Count, color = Species)) +
    geom_point(alpha = 0.8) +
    scale_color_manual(values = species_colours) + 
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x = "Country", y = "", color = "Species") + 
    theme(legend.position = "none", text = element_text(size = 20))



ggarrange(p7, p8, ncol=2, align = "h")


```

Different species were also subject to research focusing on different outcome variables (@fig-species_ebv). As for the number of studies, wild turkey (7 outcome variables) is the most well studied species in the sense of number of different outcome variables, with bobwhite (6 outcome variables) and willow ptarmigan/grouse (5 outcome variables) following closely. Research on capercaillie and black grouse have 3 outcome variables. Noticeably, two of the outcome variables, have only been the focus of two studies on grey partridge (*Perdix perdix*; effective population size) and rock ptarmigan (*Lagopus muta*; morphology).

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-species_ebv
#| fig-cap: "The distribution of EBV classes by species"

sp_data<-included_full_details |>
  separate_rows(species_if_multiple_seperate_with_a_comma_use_full_latin_names_e_g_lagopus_lagopus_lagopus_muta, sep = ", ") |>
  mutate(
    species_corrected = map_chr(
      species_if_multiple_seperate_with_a_comma_use_full_latin_names_e_g_lagopus_lagopus_lagopus_muta,
      ~ {
        res <- tryCatch(name_backbone(name = .x), error = function(e) NULL)
        if (!is.null(res) && "species" %in% names(res) && !is.na(res$species)) {
          res$species # Return corrected name
        } else {
          .x # Return original if not found
        }
      }
    )
  ) 


sp_EBV<-sp_data |> 
  separate_rows(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework, sep = ", ") |>
  group_by(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework) |>
  mutate(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework = case_when(
    outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework %in% c("Behaviour", "Display behaviour/ Gobbling", "Foraging behaviour", "Gobbling activity") ~ "Behaviour",
    TRUE ~ outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework
  )) 
  
# Aggregate data to get counts
heatmap_data <- sp_EBV |>
  mutate(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework = case_when(
    outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework %in% c("Population density *", "Species abundance", "Population growth rate") ~ "Population density/abundance/growth*",
    TRUE ~ outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework
  )) |>
  count(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework, species_corrected) |> 
  drop_na()

# Rename columns for easier use in ggplot
colnames(heatmap_data) <- c("EBVname", "Species", "Count")

heatmap_data$Species <- factor(heatmap_data$Species, levels = levels(with(species_data, reorder(Species, n))))
#species_colours
heatmap_data$colour <- species_colours[as.character(heatmap_data$Species)]


ggplot(heatmap_data, aes(x = EBVname, y = Species, size = Count, color = colour)) +
  geom_point(alpha = 0.8) +
  scale_color_manual(values = unique(heatmap_data$colour)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for readability
  labs(x = "EBV name",
       y = "Species",
       color = "Species") +  # Updated legend title
  theme(text = element_text(size = 15))


```

Only 6 studies were based the robust RCT approach, of which 5 were done on willow ptarmigan/grouse and one on wild turkeys. Four species, wild turkey, willow and rock ptarmigan and the dusky grouse (*Dendragapus obscurus*) have been studied using a before-after-control-impact (BACI) approach.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-species_design
#| fig-cap: "The design of observational studies by species"
sp_data<-included_full_details |>
  separate_rows(species_if_multiple_seperate_with_a_comma_use_full_latin_names_e_g_lagopus_lagopus_lagopus_muta, sep = ", ") |>
  mutate(
    species_corrected = map_chr(
      species_if_multiple_seperate_with_a_comma_use_full_latin_names_e_g_lagopus_lagopus_lagopus_muta,
      ~ {
        res <- tryCatch(name_backbone(name = .x), error = function(e) NULL)
        if (!is.null(res) && "species" %in% names(res) && !is.na(res$species)) {
          res$species # Return corrected name
        } else {
          .x # Return original if not found
        }
      }
    )
  ) 


sp_design<-sp_data |> 
  separate_rows(study_design_observational_studies_skip_if_simulation_study_selected_above, sep = ", ") |>
  group_by(study_design_observational_studies_skip_if_simulation_study_selected_above) #|>
  #mutate(study_design_observational_studies_skip_if_simulation_study_selected_above = case_when(
   # outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework %in% c("Behaviour", "Display behaviour/ Gobbling", "Foraging behaviour", "Gobbling activity") ~ "Behaviour",
    #TRUE ~ outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework
  #)) 
# Aggregate data to get counts
heatmap_data <- sp_design |> 
  count(study_design_observational_studies_skip_if_simulation_study_selected_above, species_corrected) |>
    mutate(study_design_observational_studies_skip_if_simulation_study_selected_above = fct_recode(study_design_observational_studies_skip_if_simulation_study_selected_above, BACI = "Before-After-Control-Impact (BACI)", RCT = "Randomized Controlled Trials (RCTs)")) |>
mutate(name = fct_relevel(study_design_observational_studies_skip_if_simulation_study_selected_above, 
                            "RCT", "BACI", "Before-After (BA)", 
                            "Control-Impact (CA)", "After (A)")) |>
  drop_na()

# Rename columns for easier use in ggplot
colnames(heatmap_data) <- c("Design", "Species", "Count")
heatmap_data$Species <- factor(heatmap_data$Species, levels = levels(with(species_data, reorder(Species, n))))
#species_colours
heatmap_data$colour <- species_colours[as.character(heatmap_data$Species)]


ggplot(heatmap_data, aes(x = Design, y = Species, size = Count, colour=colour)) +
  geom_point(alpha = 0.8)+
   scale_color_manual(values = unique(heatmap_data$colour))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for readability
  labs(x = "Study design",
       y = "Species",
       fill="") +
  theme(legend.position = "none", legend.title = element_blank()) +
  theme(text = element_text(size = 15))
```

## Summary of the literature mapping

Overall, there has been a substantial amount of research on the effects of harvest on galliform species. However, our mapping of the literature also revealed that the literature is dominated by non-experimental research and often conducted at a local or regional scale. We identified only two research projects that were based on a RCT approach, of which one of the projects (represented with five papers) were the former Norwegian project on willow ptarmigan carried out in the late 1990s. Moreover, while the willow ptarmigan (or willow grouse) is well represented in the literature (28 records), there was little research on capercaillie (4 records) and black grouse (6 records), and no experimental assessment of the effects of harvest were identified.

# Experimental design

The planned harvest experiment will rely on the existing line transect surveys gathered through [Hønsefuglportalen](https://honsefugl.nina.no/Innsyn/nb). In cooperation with Statskog and selected "Fjellstyrer" in Norway, the planned experiment will have access to a range of hunting areas from mid- and northern-Norway covering a latitudinal gradient between approx. 58 <!-- (if Njardarheim is included) --> and 69 degrees. <!--# More here -->.

In this chapter, we will provide the basis for the experimental study design for the planned hunting experiment. We will start by defining some key terms (**BOX: Glossary of terms**), and then put this into the context of the planned experiment. First we will define the harvest treatment in terms of harvest levels. Then, we will present the results from a simulation exercise that was conducted to assess the required sample sizes. Then, we will present a detailed description of how study areas (study units) were "recruited" based on a pre-determined set of criteria. Finally, a complete list of study units will be presented together with key characteristics and the randomly allocated harvest treatments.

The setup of the experiment must be carefully designed such that the potential effects of both climate and harvest on ptarmigan and forest grouse populations will be detectable in the study. This means that the experimental design must include 1) areas with enough climatic variation and 2) large enough samples, i.e., number of areas, to statistically estimate the effects of the treatments (harvest rate) and covariates (climatic variables) on survival and population growth rate. Here, we have done a power analysis using simulations, with reported values of survival and population density from the literature, to estimate the number of areas needed to get the statistical power to test the effects of experimental treatments and climate.

::: {.callout-tip appearance="minimal"}
## BOX: Glossary of terms

**General terms**

**Outcome variable:** This is the variable of interest, the one to draw inference about. Defining the outcome is part of the planning of an experiment. In our harvest experiment, the outcome variables of interest include the effects of experimental harvest on i) survival and ii) population growth.   

**Study population**: The population we will draw inference about. In strict terms, this is the willow ptarmigan population within the geographical boundaries of our study. Assuming that the willow ptarmigan population in our study area respond in a similar way to harvest mortality as other populations, we can make predictions about how the response will be in other areas. By including *covariates* that affect the outcome (response to harvest) these predictions could be made more accurate and less biased.

**Statistical population**: A statistical population is the set of subjects that we sample from and make inference from. In our case, the statistical population is the harvest areas that are included within the study area. One subject in the statistical population from which we draw some measurement is typically named a statistical unit or a sampling unit.

**Sampling frame:** The sampling frame is the “collection” of all study subjects that are available for inclusion / sampling. A structured way of thinking about this is to think of a regular grid covering the whole study area. However, in our case the grid is not regular but consists of harvest areas of different sizes. In addition, we will have to make other adjustments to the sampling frame: i) only areas with “sufficient” line transect survey data are available for inclusion, and ii) only areas larger than a cut-off size. These adjustments and their consequences should be considered before we make the final decision, as they can induce bias (see below).   

**Blocked study design**: In blocked designs, the study units are grouped (blocked) by some characteristics. In a [*randomized block design*]{.underline}, the treatment level is randomly assigned within the blocks. A block design typically means that the sample size must be larger in order to obtain the same statistical power as in non-blocked designs.    

**Internal and external validity**: The term [*internal validity*]{.underline} refers to which extent the inference drawn from the sample, e.g. in terms of cause-effect relationships are true for the sample itself. The term external validity relates to which extent one can generalize from the current sample to the larger population that we usually want to make statements about. While the former in large-part is determined by the robustness of the methods and how treatment levels are assigned etc, the latter is in large-part determined by the sampling design and to which extent the included units are representative for the population.  

**Recruitment/inclusion of study subjects**

The rules or algorithms chosen when recruiting / including study subjects. Since we cannot usually measure all subjects in the population, this usually involves some form of probabilistic or non-probabilistic sampling. Should be constructed to minimize the potential for *biased* estimates of inference. For instance, if all included subjects share a common characteristic that is not shared with other members of the statistical population, then the results might be biased. This is termed selection bias. Selection bias occurs whenever the subjects that are included in the study systematically deviate from the study population. Note that selection bias might also apply when assigning treatment level to a study unit in an experiment (covered below).  

**Probabilistic sampling to recruit study subjects:** This involves using recruiting units into the study by some sort of random (probabilistic) sampling. In the basic form (i.e. [*random sampling*]{.underline}), all units have the same probability of being recruited. In [*stratified random sampling*]{.underline}, the units are sorted along one or more covariates to ensure all values along the covariate are represented with a certain proportion in the sample. Under proportional stratification, all “strata” are represented with a proportion that is similar to their proportion in the statistical population. Under balanced stratification, the number of recruited units are balanced (e.g. similar) across strata. Under [*restricted random sampling*]{.underline}, the initial sampling might be random but it is combined with rules that make the next decisions non-probabilistic. Examples include “do not include units that are located next to each other” etc. Another sampling strategy that combines probabilistic and non-probabilistic sampling is [*systematic sampling*]{.underline} when the selection of the first unit is decided by a random draw.

**Non-probabilistic selection to recruit study subjects:** These approaches include methods that use non-probabilistic criteria to select study subjects. Such criteria could include geography, certain characteristics (i.e - placement along a covariate axis but with no random selection – as opposed to stratified random sampling) and expert knowledge. A very specific type of non-probabilistic selection is [*convenience sampling*]{.underline} where the units that for some reason are “easiest” to include are selected. One method that further resembles “stratified random sampling” is what is often termed [*proportional quota sampling*]{.underline} in the medical literature. Here, the number of units that are included are proportional to the size of the strata, but the selection of units to include are not done at random. In cases when the total size of the population from which to draw sampling units from is unknown, this method will be known as [*non-proportional quota sampling*]{.underline}. Finally, *purposive sampling* is a rather broad class of approaches where each study unit is recruited into the study based on one or more specific characteristics. In strict terms, a non-probabilistic selection of study subjects does not rely on a sampling frame.   

**Assignment of treatment level to each study unit**  

In manipulative experiments, there also need to be rules for assigning treatment level to the study subjects. This can be done by [*random assignment*]{.underline} or non-random ([*quasi random*]{.underline}) assignment. Under [*random assignment*]{.underline} of experimental treatment level the treatment level (or control vs treatment) is randomly assigned to each unit. This is efficient in terms of assuring internal validity, as there will be no confounding effects from covariates. In [*blocked designs*]{.underline}, treatment levels are randomly assigned within blocks. In a [*crossover study*]{.underline} or trial, each study subject obtains a series of treatments allocated at random. In most cases, the study is designed so that all included subjects receive all treatments in a random order. This will result in a [*balanced study design*]{.underline}, where the number of observations for each treatment are the same. This is in contrast to parallel studies or non-crossover studies where each unit is exposed to only one treatment level. For heterogeneous samples, one advantage (depending on the analyses methods) is that each study subject can serve as their own control.

**Key references**

@christie2019simple and @morrison2008wildlife were used to develop the above text.
:::

## Experimental harvest treatment

In the planned experiment, we will base the harvest treatment for willow ptarmigan on the same experimental levels as the former large-scale experiment carried out in the 1990s in Norway [@Sandercock2011]: 0%, 15% and 30% harvest, respectively. This will make our results directly comparable to those results. For black grouse and capercaille, we will apply 0%, 10% and 20% harvest. The rationale for this is that these birds have a slower life history and higher natural survival rates, and thus we expect that harvest is additive for these species.

### Random allocation of harvest levels

Harvest treatment for each study unit will be randomly allocated based on a crossover study design. As the planned experiment will have three experimental harvest levels (see above) and last for three seasons, one study unit, i.e., one hunting area, will be subject to each harvest treatment. The sequence of the treatments will be randomly decided for each study unit. To ensure a balanced design in terms of the combined allocation of harvest levels across years we will make a balanced set of sequences and randomly allocate a sequence to a study unit. Since there are three harvest treatments there will be six unique sequences of treatments of which each area will be given for the duration of the experiment. For example, if there would be 30 study units included in the experiment, each treatment sequence would be given for $30/6 = 5$ areas.

### Outcome variables

Based on the distance sampling line transect survey data collected from the study units, we will focus on the effects of harvest on two outcome variables, namely i) (annual) survival probability under different harvest treatments and ii) population growth under different harvest treatments. In addition to including information about harvest treatment, we will include information about climatic and biotic factors (in particular rodent dynamics) known to affect grouse population dynamics or is expected to affect the effect of harvest on the outcome variables of interest.

We will base our inference on the model presented by @nilsen2023 to assess effects on survival. This model allow us to directly assess effects of harvest on survival rates, as it includes information about the population structure to estimate survival and recruitment rates separately. In this model, population growth rate is not estimated directly but is a derived parameter; thus we will use a modelling framework similar to that presented in (@Bowler2020) to investigate effects directly on population growth.

### Estimation of hunting quota

Each year, prior to the hunting season, a hunting quota that corresponds to the randomly allocated harvest level will be allocated based on the annual line transect surveys. The population density in a given hunting area is estimated using distance sampling techniques (@nilsen2023, @Bowler2020). The transect surveys are coordinated through [Hønsefuglportalen](https://honsefugl.nina.no/Innsyn/nb) and the results can be used to estimate the total population density as well as the number of adults and juveniles (that years production of chicks). In most hunting areas the yearly production is an important parameter in setting the yearly quotas, but in the harvest experiment the quotas are based on the total population size. For example, if the harvest level for a given year is 30 % and the density is estimated as 15 birds/km^2^ in a area of 100 km^2^, that years hunting quota will be $density \cdot area \cdot harvest level = 15 \cdot 100 \cdot 0.3 = 450$ birds.

## Simulation to assess sample size

### Population parameters

The willow ptarmigan populations that will be used in the harvest experiment are all surveyed approximately one month after breeding when the chicks are large enough to take flight. To simulate the effects of harvest and climate on willow ptarmigan survival and population growth rate, we used a post breeding projection matrix ($A_t$) with two age classes: juveniles (0-1 year) and adults (1+ years):

$$
A_t =
\begin{bmatrix}
f_{juv,t} & f_{ad,t} \\
s_{juv,t} & s_{ad,t}\\
\end{bmatrix}
$$ where, $s_{juv,t}$ and $s_{ad,t}$ are age-specific survival rates and $f_{juv,t}$ and $f_{ad,t}$ are recruitment (number of juveniles produced per female) multiplied with age-specific survival. A projection matrix contains both survival and reproduction for a population and can be used to calculate the population size for the next time-step, *t + 1*, directly from the age-specific population size at the last time step, *t*:

$$
N_{t+1} = A_tN_t
$$

where *A* is the projection matrix and $N_t$ is a vector containing age-specific populations sizes at time *t*. In these simulations we assumed equal recruitment for the two age classes. Survival estimates were collected from Sandercock et al., (2011) while recruitment was modelled as a normally distributed variable with an average of two juveniles per female and SD = 0.3. The yearly population growth rate, $r_t$, was calculated as:

$$
r_t = log(N_{juv, t+1} + N_{ad, t+1}) - log(N_{juv, t} + N_{ad, t})
$$

where $N_{juv}$ and $N_{ad}$ is the number of juveniles and adults at time *t* and *t+1*. Both survival rates and population sizes were simulated with environmental stochasticity, i.e., year-to-year variation, and observation error to better reflect the **real** parameters that will be collected during the harvest experiment.

### Harvest treatment and covariate

To evaluate and analyse the effects of harvest, we assigned the simulated hunting areas with a harvest treatment, i.e., a harvest rate of either 0, 15 % or 30 % each year over the study period of three years. Because we aim to test the effects of all treatments in all areas, the treatments were randomly given to each hunting area without replacement to ensure that each area cannot have the same treatment twice. In addition to harvest treatment, we simulated the effect of a climate covariate directly on survival. The “climatic variable” chosen was latitude and it was simulated with a slope of 0.003. This does not mean that we expect to see any direct effect of latitude in the real experiment, but rather it was included as a proof-of-concept to test the effects of a climatic covariate. To investigate how many areas are needed in the harvest experiment to be able to estimate the effects of both harvest treatments and climate we simulated ptarmigan populations over a range of hunting areas, from 15 to 35. For each number of hunting areas, we did 500 simulations over three years.

### Results

#### Survival

The effects of harvest treatment and the climatic variable on survival was modelled using a generalized linear model with a quasibinomial error function. In the simulated harvest experiment, both harvest treatment had a clear effect on survival which was evident for all sample sizes (number of areas) for the 30 % harvest, while the effect of 15 % harvest was found in \> 95 % of the simulations from 17 areas and upwards. The effect of the climatic variable was found in \> 84 % of the simulations from 20 areas and upwards.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-survival_sim
#| fig-cap: "The proportion of simulations (n = 500), over a range of hunting areas, that has a significant effect of harvest treatment (15 and 30 %) and a climatic variable on survival."

results_s <- readRDS("data/lirype_survival.RDS")
  
## Plotting proportion of simulations with p-values above 0.95 - Survival

## Changing names for facet_wrap
facet_labels <- c(
  "harvest_rate0.15" = "Harvest rate = 15 %",
  "harvest_rate0.3" = "Harvest rate = 30 %",
  "latitude" = "Climatic covariate"
)

lirype_s <- results_s %>% 
  filter(term != "(Intercept)") %>% 
  ggplot(aes(x = n_areas, y = prop_p)) + 
  geom_line(linewidth = 1.3)+
  geom_hline(yintercept = 0.95, color = "red", linewidth = 1.2)+
  facet_wrap(~term, labeller = as_labeller(facet_labels))+
  theme_minimal(base_size = 15)+
  labs(y = "Proportion of simulations with significant effect",
       x = "Number of hunting areas in the simulation",
       title = "Effects of hunting on survival")
lirype_s

```

#### Population growth rate

The effects of harvest treatment and the climatic variable on the population growth rate was modelled using a linear regression. The 30 % harvest treatment had an effect on the population growth rate in \> 95 % of the simulations where the number of areas was above 21. The effect of 15 % harvest was harder to detect and it had an effect in about 50 % of the simulations with the number of areas above 28. The effect of the climatic variable on the population growth rate was found in \> 80 % of the simulations from 27 areas and upwards.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-survival_r
#| fig-cap: "The proportion of simulations (n = 500), over a range of hunting areas, that has a significant effect of harvest treatment (15 and 30 %) and a climatic variable on the population growth rate"

results_r <- readRDS("data/lirype_growth_rate.RDS")
  
## Plotting proportion of simulations with p-values above 0.95 - population growth rate

## Changing names for facet_wrap
facet_labels <- c(
  "harvest_rate0.15" = "Harvest rate = 15 %",
  "harvest_rate0.3" = "Harvest rate = 30 %",
  "latitude" = "Climatic covariate"
)

lirype_r <- results_r %>% 
  filter(term != "(Intercept)") %>% 
  ggplot(aes(x = n_areas, y = prop_p)) + 
  geom_line(linewidth = 1.3)+
  geom_hline(yintercept = 0.95, color = "red", linewidth = 1.2)+
  facet_wrap(~term, labeller = as_labeller(facet_labels))+
  theme_minimal(base_size = 15)+
  labs(y = "Proportion of simulations with significant effect",
       x = "Number of hunting areas in the simulation",
       title = "Effects of hunting on the population growth rate")
lirype_r

```
