---
title: "Experimental harvest of ptarmigan, black grouse and capercaillie:" 
authors: 
  - name: Erlend B. Nilsen
    address: NINA Trondheim
    email: erlend.nilsen@nina.no
  - name: Lasse Frost Eriksen
    address: NINA Trondheim
    email: lasse.eriksen@nina.no
  - name: Christoffer Høyvik Hilde
    address: NINA Trondheim
    email: christoffer.hilde@nina.no
  - name: Matthew Grainger
    address: NINA Trondheim
    email: matthew.grainger@nina.no
format: html
bibliography: References_SysMap_Hunting_Galliforems.bib
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

require(NinaR)
require(xtable)
require(knitr)
require(tidyverse)
require(ggpubr)
library(rgbif)
library(purrr)


```

# Introduction {#inl}

-   In most regions of the world, harvest of wild bird populations for food and recreation has been common throughout human history (refs). Along with the development of wildlife management as a discipline, research that aim to facilitate sustainable use of wild bird populations has prevailed in the scientific and popular literature (refs).

-   Galliformes are popular game species throughout their distributional range. Out of a total of XX speceis, XX are considered game species. Research on galliform harvest has a long tradion, manifest through focus on iconic game species such as ptarmigan, grouse, sage grouse etc. But little is known about how frequent the different species have been studied, and where.

-   Recreational harvest can interfere with several aspects of a species ecology. Most obviously, survival rates, but also population growth rates, etc. EBV was developed to synthesise ecological knowledge, and easily lends itself.

-   

Sustainable harvest management of our game resources is a key goal in Norwegian nature management. In order to achieve sustainable management, it is generally recommended to use an adaptive approach, where one actively searches for new knowledge that is necessary to fill existing knowledge gaps [@Johnson1997]. In the last couple of decades a series of environmental challenges have arisen, which are expected to largely affect hunting management. These challenges are largely linked to climate change and other man-made changes in the natural environment, such as habitat loss and degradation, but also changes in other components of the socio-ecological systems wildlife species are part of and which are expected to influence how sustainable management is defined [@Bowler2020; @Eriksen2018; @Henden2020; @Mansson2023]. With increased focus in society on the loss of biological diversity on a global and national scale, one can further anticipate increased expectations in society that recreational hunting is carried out in a sustainable way with minimal risk of overharvesting. This can eventually be assumed to lead to stricter requirements for documentation of sustainability. More recent conceptual studies of adaptive management have discussed how these new situations not only force new knowledge needs, but that it can also lead to changed objectives and measurement parameters for adaptive management [@Mansson2023]. <!--# Perhaps give some specific examples at the end of the sentence ? -->In the light of the drastic changes happening across the world's ecosystems, largely driven by anthropogenic activity, it is vital to simultaneously investigate and quantify the effects of harvest and environmental variability on wildlife populations.

A central concept for predicting the effects of hunting on wildlife populations is the relationship between additive hunting mortality (which comes in addition to other mortality) vs compensatory hunting mortality (which is compensated by other mortality being reduced in proportion to the harvest mortality - see @Perone2013 and @Sandercock2011). Our knowledge of this comes largely from experimental studies carried out at the end of the 1990s under the auspices of the "Rypejaktprosjektet" (with Hans Chr. Pedersen, NINA, as project leader). Here it was concluded that hunting mortality was partially compensatory at low harvest rates, while it was additive or even super-additive at high harvest rates [@Sandercock2011; @Pedersen2004]. This knowledge base was produced by carrying out large-scale experimental studies, where the harvest treatments (0%, 15% and 30%, respectively) were distributed randomly between the study areas. Effects were investigated both directly in terms of survival (using radio-tagged birds - see @Sandercock2011) and indirectly via population densities (see @Pedersen2004). This project focused exclusively on willow ptarmigan (*Lagopus l. lagopus*), and no similar knowledge exists for e.g. black grouse *(Lyrurus tetrix*) or capercaillie (*Tetrao urogallus*).

Whether results from the former large scale experiment is transferable in time (i.e. to the current environmental conditions) and to other species (i.e. black grouse and capercaillie) is an empirical question that cannot be answered directly by existing data. For example, one can expect that climate change will have significant effects on the trophic interactions that largely affect grouse population dynamics and densities [@Bowler2020, @Henden2020]. Changes in these interactions will have the potential to directly affect grouse populations' potential to compensate for hunting mortality (see discussions in e.g. @Perone2013). Similarly, large changes in the size of the grouse population could have similar effects, since the degree of compensation can be closely linked to density dependence [@Sinclair1996]. One way to address these uncertainties would be to conduct a conceptual replication of the experiments carried out in the 1990s. Such conceptual replication studies (*sensu* @Nakagawa2015) are rare in ecology [@Fraser2020], but have recently been strongly recommended to examine the validity and generalisability of previously conducted studies [@Fraser2020; @Nakagawa2015].

However, several aspects should be carefully investigated before carrying out such a replication. Initially, it is a prerequisite to conduct an up-to-date mapping of the research-based literature on the sustainable management of galliformes from a climate change perspective (see also @Grainger2020), to identify gaps in the literature and to ensure that the replication rests on the best available knowledge. Further, it requires thorough planning of how best to design such an experiment in a way that sample sizes and climatic gradients are sufficiently estimated in order to close the most pressing knowledge gaps. This report is a central part of the planning stage of such a project, where the intent is to apply experimental harvest treatments to populations of willow ptarmigan, black grouse and capercaillie in Norway. The project is planned to take place across a three-year experiment within a select number of hunting areas managed by Statskog across Norway (from Agder in South to Troms in the North). The report includes two main sections:

-   A systematic mapping of the literature related to effects of harvest on galliformes

-   A thorough evaluation of the experimental design including both the sampling of study subjects and assignment of treatment levels

We will first present the results from the systematic mapping of the literature, and then the evaluation of experimental design. In the latter chapter, we will present a detailed description of the suggested design for the new harvest experiment.

# Systematic mapping of the literature

This systematic map is produced as part of the preparation of a large-scale harvest experiment in Norway, as outlined above. We aimed to identify studies that have quantitatively assessed the effects of legal harvest (i.e., managed hunting) on galliformes, to ensure that the planned experimental studies on the sustainability of galliform hunting are based on the best available evidence. This mapping aims to identify, categorise, and synthesise existing research on how hunting mortality interacts with ecological and environmental factors. By systematically compiling relevant studies, we provide a foundation for assessing knowledge gaps and determining the generalisability of past research.

The [protocol](https://osf.io/ny2jt/) i.e., the workflow of identifying, categorising and synthesising the exisiting literature, was prepared prior to conducting the literature searches. Here, we include the relevant information and report any deviation from the protocol.

## Methods

### Search for articles

We used a PIO (Population, Intervention, Outcome) question structure to help develop the search strategy.

-   Population (P): Bird species within the order Galliformes

-   Intervention (I): Legal (managed) hunting or harvest

-   Outcome (O): The ecological impacts of managed hunting on population dynamics, demography, genetics or behaviour 

We searched for academic literature (publications) in two search platforms; Web of Science and lens.org. Specifically, we searched the Web of Science Core Collection (WOS.SCI: 1987 to 2025, WOS.AHCI: 1987 to 2025, WOS.ESCI: 2020 to 2025 and WOS.SSCI: 1987 to 2025) using the NINA library subscription and the following search string:

((TS=(galliform\* OR grouse OR ptarmigan OR Phasianidae OR pheasant OR "lagopus lagopus" OR "lagopus muta" OR "wild turkey" OR "bobwhite" OR partridge))

AND (TS=("hunt" OR "harvest")))

AND SU=("Ecology" OR "Wildlife" OR "Environment")

For lens.org we used the following search string:

(abstract:("galliform" OR "grouse" OR "ptarmigan" OR "Phasianidae" OR "pheasant" OR "lagopus lagopus" OR "lagopus muta" OR "wild turkey" OR "bobwhite" OR "partridge") OR title:("galliform" OR "grouse" OR "ptarmigan" OR "Phasianidae" OR "pheasant" OR "lagopus lagopus" OR "lagopus muta" OR "wild turkey" OR "bobwhite" OR "partridge")) AND abstract:("hunt" OR "harvest") OR title:("hunt" OR "harvest") AND field_of_study:("Ecology" OR "Wildlife" OR "Environment")

Adding "capercaillie", "lyrurus" and "tetrao" did not result in additional eligible papers. In addition, a backward and forward citation chasing (using CitationChaser) was performed based on the three first benchmark papers listed below [@Sandercock2011; @Pedersen2004; @Sedinger2005].

### Benchmark papers

The following benchmark papers were used to assess the effectiveness of the search string. We expected to find all these papers in a review of this question. We recorded if each benchmark paper was found in the searches (Wo and lens.org).

-   Sandercock, B. K., E. B. Nilsen, H. Brøseth, and H. C. Pedersen. 2011. Is hunting mortality additive or compensatory to natural mortality? Effects of experimental harvest on the survival and cause-specific mortality of willow ptarmigan. Journal of Animal Ecology 80:244-258.

-   Pedersen, H. C., H. Steen, L. Kastdalen, H. Brøseth, R. A. Ims, W. Svendsen, and N. G. Yoccoz. 2004. Weak compensation of harvest despite strong density-dependent growth in willow ptarmigan. Proceedings of the Royal Society of London Series B-Biological Sciences 271:381-385.

-   Sedinger, J.S. & Rotella, J.J. 2005. Effect of harvest on sage-grouse Centrocercus urophasianus populations: what can we learn from the current data? Wildlife Biology, 11, 371–375.

-   Pedersen, Å. Ø., E. M. Soininen, S. Unander, M. H. Willebrand, and E. Fuglei. 2014. Experimental harvest reveals the importance of territoriality in limiting the breeding population of Svalbard rock ptarmigan. European Journal of Wildlife Research 60:201-212.

-   Brøseth, H., J. Tufto, H. C. Pedersen, H. Steen, and L. Kastdalen. 2005. Dispersal patterns in a harvested willow ptarmigan population. Journal of Applied Ecology 42:453-459.

-   Brøseth, H., E. B. Nilsen, and H. C. Pedersen. 2012. Temporal quota corrections based on timing of harvest in a small game species. European Journal of Wildlife Research 58:797-802.

-   Brøseth, H., and H. C. Pedersen. 2000. Hunting effort and game vulnerability studies on a small scale: a new technique combining radio-telemetry, GPS and GIS. Journal of Applied Ecology 37:182-190. 

-   Henden, J.-A., R. A. Ims, N. G. Yoccoz, E. J. Asbjørnsen, A. Stien, J. P. Mellard, T. Tveraa, F. Marolla, and J. U. Jepsen. 2020. End-user involvement to improve predictions and management of populations with complex dynamics and multiple drivers. Ecological Applications n/a:e02120. 

-   Small, R.J., Holzwart, J.C. & Rusch, D.H. 1991. Predation and hunting mortality of ruffed grouse in central Wisconsin. Journal of Wildlife Management, 55, 512–520. 

In our searches, all benchmark papers were found. @Henden2020 and @small1991 were not indexed in lens.org but were found in Web of Science.

### Article screening

We used the online screening tool Rayyan [@ouzzani2016rayyan] for screening the results of the article search. A pre-determined inclusion/exclusion criteria was used when screening eligible studies. Excluded papers will be shown in a flowchart using the ROSES flow chart [@Haddaway2018]. Included studies was de-duplicated based on DOI and title matches using the deduplication tool in Rayyan.

After de-duplication, title and abstract screening was performed using the eligibility criteria. If uncertainty remained, the full text version was visited to confirm/reject. A set of 54 papers was reviewed by three members of the review team to ensure consistency.

### Eligibility criteria

We reviewed scientific articles from all parts of the world covering all galliform species. Studies of both native and introduced populations were eligible. Studies of all spatial scales and across all relevant ecosystems were considered eligible. Only studies documenting the effects of legal harvest on galliform populations were eligible. Studies of poaching were not eligible. The effects must have been assessed quantitatively for the study to be eligible. Eligible outcomes included effects of harvest on population abundance, density, sex- and age structure, recruitment, survival, behaviour, genetics, physiology and distribution. Studies focusing on heavy metal (e.g. lead) poisoning or content in wild birds from shooting were not eligible. Both observational and experimental study designs were eligible. Mathematical simulation studies were eligible.

### Title and abstract screening stage

We used a decision tree to make decisions about inclusion and exclusion at the *Title* and *Abstract* stage. The tree is presented in the protocol and in the *Appendix* of this report. The decision tree helped the reviewers to make decisions about eligibility of each paper in a structured way. Dual screening was carried out on 20% of papers (n = 183). Agreement between reviewers was assessed using the kappa statistic which was over 0.7, which is deemed acceptable.

### Full text eligibility

Papers included at the *Title* and *Abstract* stage was taken forward to the full text screening stage. An RIS file of included articles was downloaded from Rayyan and added to a Zotero database. PDF versions of each paper were searched using Zotero’s “Find Available PDF” tool, ResearchGate, Google Scholar and Google. If a PDF was not available online (or was not open to us), the paper was excluded at this stage of the review. Full text papers were excluded if they did not meet the following criteria:

-   Written in English, Norwegian, Swedish or Danish
-   Focused on species from the order Galliformes
-   Documented the effect of legal or managed hunting on galliformes
-   Assessed the effect of hunting quantitatively

### Data coding

From all papers that met the eligibility criteria, we extracted key variables (a.k.a. “data coding”). In addition to metadata about the publication (publication year, journal name etc.) we extracted from the full-text information about study species, study area location and extent, research approach and experimental design, outcome variables and to which extent the paper reported harvest rates and to which extent the study assessed whether harvest was additive or compensatory. For outcome variables, we used the Essential Biodiversity Variable (EBV) framework [@pereira2013essential; @navarro2017monitoring], and coded both for EBV class (high order classification) and EBV name (lower order classification). A detailed description of data coding is presented in the protocol and in the *Appendix*. 

## Results from the literature review

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

library(googlesheets4)
# Read data
#data <- read_sheet("https://docs.google.com/spreadsheets/d/1Shdb7vfLn-bnZd0-FT2Ngqovxi1iEn-MYLnTtBPpBQ4/edit?resourcekey=&gid=968606953#gid=968606953")
#

#saveRDS(data,"data/GS_data_Raw.RDS")
data <- readRDS("data/GS_data_Raw.RDS")



data<-data |> 
  distinct(`Study ID`, .keep_all = TRUE)

excluded <- data |>
  filter(!rowSums(across(4:8, ~ . == "Yes")) == 5)
included <- anti_join(data, excluded)
```

Our searches returned 1023 articles in total, with 542 from Web of Science, 112 from lens.org, 9 benchmark articles and 192 from backward-, and 168 from forward citation chasing from the three benchmark articles mentioned above (i.e., @Sandercock2011; @Pedersen2004; @Sedinger2005). Of these, 173 were duplicates which were removed from the database. We screened the remaining 850 articles for eligibility at the *Abstract* and *Title* stage. Of the 850 included articles, a total of 200 were moved further to full-text screening, of which 166 papers that had an "include" decision and 34 a "maybe" decision. There were 9 conflicts (where reviewers disagreed on eligibility). All these papers were taken through to the fulltext stage and assessed for eligibility. Exclusion decisions (650 records) were due to the research not focusing on the effects of legal or managed hunting (365 records), not being focused on galliformes (276 records), not quantitative (23 records), abstract not found online (16 records), or abstract not available in English, Norwegian, Swedish or Danish (3 records).

At the fulltext stage `r dim(excluded)[1]` articles were excluded (3 with no available pdf, 6 not focused on galliformes, 12 not assessing the effects of hunting quantitatively and the rest, 61, not documenting the effects of legal harvest). This left `r dim(included)[1]` papers that went forward to the data extraction stage.

![ROSES Flowchart showing the number of papers included or excluded in the review](images/roses%20(3)-01.png){#fig-prisma}

```{r}
#| echo: false
#| warning: false
#| message: false

library(synthesisr)
all_RIS <- read_refs("data/articles.ris")

included_full_details <- left_join(included, all_RIS, by = c("Study ID" = "accession_zr"))

included_full_details <- included_full_details |> janitor::clean_names()
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: colour palette 
species_colours <- c(
  "Centrocercus urophasianus" = "#736F4C",  # Sage-grouse (greyish-brown)
  "Dendragapus obscurus" = "#5A4A42",       # Dusky Grouse (brown-grey)
  "Lagopus lagopus" = "#85361C",            # Willow Ptarmigan (reddish-brown)
  "Lyrurus tetrix" = "#1C1C1C",             # Black Grouse (black)
  "Tetrao urogallus" = "#2B2B2B",           # Capercaillie (dark grey/black)
  "Tetrastes bonasia" = "#8C7051",          # Hazel Grouse (brownish-grey)
  "Perdix perdix" = "#A67B5B",              # Grey Partridge (grey-brown)
 "Lagopus muta" = "#6E6259",               # Rock Ptarmigan (darker grey-brown)
  "Alectoris graeca" = "#B0A48B",           # Rock Partridge (grey-brown)
  "Alectoris rufa" = "#A5694F",             # Red-legged Partridge (red-brown)
  "Alectoris chukar" = "#B48B7D",           # Chukar Partridge (beige-brown)
  "Bonasa umbellus" = "#7A5F47",            # Ruffed Grouse (brown-grey)
  "Colinus virginianus" = "#9E8358",        # Northern Bobwhite (tan-brown)
  "Meleagris gallopavo" = "#4A2D1D",        # Wild Turkey (dark brown/bronze)
  "Phasianus colchicus" = "#B22222"         # Common Pheasant (red and green)
)

```

Legal hunting of galliformes has been subject to extensive research. In our searches we identified studies published as far back as 1963, but we are aware that research on this topic was carried out even before that, however these studies were not covered by our searches.<!--# Because we didnt search further back in time or because the keywords didnt match ? If the former- we should explain why we limited the search to a certain time-period  --> The number of studies per year peaked in 2020 (8 studies) and the long-term average was 2.93 studies per year (@fig-Publication_year).

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-Publication_year
#| fig-cap: "The number of studies per year"

included_full_details |>
  group_by(year) |>
  tally() |>
  drop_na() |>
  ggplot(aes(year, n)) +
  geom_histogram(stat = "identity", fill=species_colours[3]) + # colour of a Willow Ptarmigan
  labs(
    x = "Year",
    y = "Number of records",
    #fill = "Response"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        text = element_text(size =13))
```

The most common geographic location for studies was in the USA (99 studies), followed by Norway (16) and the UK (9) (@fig-location). In the USA studies were carried out in 33 States (Alaska, California, Colorado, Florida, Georgia, Idaho, Illinois, Indiana, Kansas, Kentucky, Louisiana, Maine, Michigan, Mississippi, Missouri, Montana, Nevada, New York, North Carolina, Ohio, Oklahoma, Oregon, Pennsylvania, Poland, Rhode Island, South Carolina, South Dakota, Tennessee, Texas,Utah,Virginia, West Virginia, Wisconsin). <!--# a map showing the studies would be cool, maybe with the world's distribution of harvested galliforms, or just galliforms in general, as a layer (If that layer exists)? -->

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-location
#| fig-cap: "Geographic location of included studies"

included_full_details |>
  separate_rows(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma, sep = ", ") |> # Split at comma and create new rows
  group_by(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma) |>
  mutate(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma = case_when(
    study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma %in% c(
      "Alaska", "California", "Colorado", "Florida", "Georgia","Georiga", "Georgia (USA)", "Idaho", "Illinois", "Indiana", "Kansas", "Kentucky", "Louisiana", "Maine", "Michigan", "Mississippi", "Missouri", "Montana", "Nevada", "New York", "North Carolina", "Ohio",
      "Oklahoma", "Oregon", "Pennsylvania", "Poland", "Rhode Island", "Rhode island", "South Caolina", "South Caraolina", "South Carolina", "South Dakota", "Tennessee", "Texas", "Texas,Texas", "US", "Utah",
      "Virgina", "Virginia", "West Virginia", "Wisconsin", "Wisonsin"
    ) ~ "USA",
    TRUE ~ study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma
  )) |> 
  mutate(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma= case_when(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma %in% c("United Kingdom", "UK", "Scotland", "England")~"UK",
            TRUE~study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma)) |> 
tally() |>
  drop_na() |>
  ggplot(aes(reorder(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma, n), n)) +
  geom_histogram(stat = "identity", fill=species_colours[3]) +
  labs(
    x = "Study area location",
    y = "Number of records"
  ) +
  #ggthemes::theme_base() +
  theme_minimal()+
  coord_flip() +
  theme(text = element_text(size = 13))
```

The extent of the study areas were predominately regional or local, with very few studies at the site level (@fig-Study_area). Few studies were of national or multinational scale. Following our definition used here, regional studies typically cover one mountain area or a county, whereas a local study would cover a municipality or parts of a municipality. Note that this contrasts previous calls for more large-scale research [@estes_spatial_2018] that might be needed to uncover the ecological dynamics and to provide input to natural resource management policy at larger spatial scales. <!--#  Not needed here, but if this is going to be published we could consider classifying the extent of study area by using km^2, study done on jungle fowl in the Vatican would be considered a national area using our current methodology but might still be smaller than the majority of regional areas in the USA for example. Bad example but you get my point :) -->

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-Study_area
#| fig-cap: "The study area extent"

included_full_details |>
  group_by(study_extent_size_of_the_study_area) |>
  tally() |>
  drop_na() |>
  #ggplot(aes(reorder(study_extent_size_of_the_study_area, n), n)) +
  mutate(study_extent_size_of_the_study_area = fct_recode(study_extent_size_of_the_study_area,
             Regional = "Regional (e.g. a county, a larger mountain area etc)", 
             Local = "Local (e.g. a municipality or part of municipality)", 
             `Site level` = "Site level (< 1km2)")) |>
  mutate(name = fct_relevel(study_extent_size_of_the_study_area, 
                            "Multinational", "National", "Regional", "Local", "Site level")) |>
  ggplot(aes(x=name, y=n)) +
  geom_histogram(stat = "identity", fill="#85361C") +
  labs(
    x = "Spatial extent",
    y = "Number of records"
  ) +
  theme_minimal() +
  coord_flip() +
  theme(text = element_text(size = 15))

```

Most studies covered in our mapping were observational (n = 106), but there were also a substantial number of studies that relied on simulations or scenario analysis (n = 35; @fig-Research_design a)). The sum of observational studies based on empirical data and studies based on simulations is higher than the total of 129 papers that were found eligible. This is because some studies used both research approaches and thus were included in both categories. In agreement with previous assessments of study designs in ecological research <!--# Add reference! -->most of the studies that presented empirical data were based on simple (and less robust) study designs (@fig-Research_design b)). There were only 6 papers using a "Randomized Controlled Trial (RCT)" approach, the study design that leads to the highest level of evidence for causal understanding [@christie2019simple] and these were limited to Norway (5) and the USA (1). The Norwegian articles were all from the same research project/experiment conducted in the late 1990s that was mentioned above.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-Research_design
#| fig-cap: "Type of research carried out"

p1 <- included_full_details |>
  separate_rows(research_type, sep = ", ") |> # Split at comma and create new rows
  group_by(research_type) |>
  tally() |>
  drop_na() |>
  mutate(research_type = fct_recode(research_type, `Empirical data` = "Observational / analysis of empirical data", `Simulation based` = "Simulation based /Scenarios")) |>
  ggplot(aes(reorder(research_type, n), n)) +
  geom_histogram(stat = "identity", fill=species_colours[2]) +
  labs(
    x = "Research type",
    y = ""
  ) +
  theme_minimal() +
  coord_flip() +
  theme(text = element_text(size = 15))


p2 <- included_full_details |>
  group_by(study_design_observational_studies_skip_if_simulation_study_selected_above) |>
  tally() |>
  drop_na() |>
  mutate(study_design_observational_studies_skip_if_simulation_study_selected_above = fct_recode(study_design_observational_studies_skip_if_simulation_study_selected_above, BACI = "Before-After-Control-Impact (BACI)", RCT = "Randomized Controlled Trials (RCTs)")) |>
mutate(name = fct_relevel(study_design_observational_studies_skip_if_simulation_study_selected_above, 
                            "RCT", "BACI", "Before-After (BA)", 
                            "Control-Impact (CA)", "After (A)")) |>
  ggplot(aes(x=name, y=n)) +
  geom_histogram(stat = "identity", fill=species_colours[3]) +
  labs(
    x = "Study design",
    y = "Number of Responses"
  ) +
  theme_minimal() +
  coord_flip() +
  theme(
    text = element_text(size = 15),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, margin = margin(t = 10)),  # Adjust rotation and margin
    axis.title.x = element_text(margin = margin(t = 15))  # Add space for axis title
  )

ggarrange(p1, p2, ncol = 1, labels = c("a)", "b)"), 
          align= "v")



```

```{r}
#| echo: false
#| eval: false
#| warning: false
#| message: false
#| label: fig-Study_design_by_country
#| fig-cap: "Study design by country"

country_data<-included_full_details |>
  separate_rows(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma, sep = ", ") |> # Split at comma and create new rows
  group_by(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma) |>
  mutate(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma = case_when(
    study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma %in% c(
      "Alaska", "California", "Colorado", "Florida", "Georgia","Georiga", "Georgia (USA)", "Idaho", "Illinois", "Indiana", "Kansas", "Kentucky", "Louisiana", "Maine", "Michigan", "Mississippi", "Missouri", "Montana", "Nevada", "New York", "North Carolina", "Ohio",
      "Oklahoma", "Oregon", "Pennsylvania", "Poland", "Rhode Island", "Rhode island", "South Caolina", "South Caraolina", "South Carolina", "South Dakota", "Tennessee", "Texas", "Texas,Texas", "US", "Utah",
      "Virgina", "Virginia", "West Virginia", "Wisconsin", "Wisonsin"
    ) ~ "USA",
    TRUE ~ study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma
  ))

# Aggregate data to get counts
heatmap_data <- country_data %>%
  count(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma, study_design_observational_studies_skip_if_simulation_study_selected_above)|> 
  drop_na()

# Rename columns for easier use in ggplot
colnames(heatmap_data) <- c("Country", "Study_design", "Count")
ggplot(heatmap_data, aes(x = Country, y = Study_design, fill = Study_design)) +
  geom_tile(aes(alpha=Count)) +
 #scale_fill_manual(values = species_colours) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for readability
  labs(x = "Country",
       y = "Study design",
       fill = "")
```

We also extracted information about harvest rates (or percentage of the population removed by harvest) in 71 papers. The remaining papers did not explicitly report such information (@fig-Harvest_rateAdditive a)). Moreover, among the papers that reported harvest rate or harvest mortality, only 27 made any assessment to which extent harvest was additive or not (@fig-Harvest_rateAdditive b)). Note however that this is somewhat ambiguous, and it might depend on how the concept of additivity is defined: although we did not quantify this by direct coding we note that much fewer studies were in fact able to properly document additivity vs compensation via a robust quantitative approach<!--# removed for now: (See also discussion in ref) -->.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-Harvest_rateAdditive
#| fig-cap: "Reporting of harvest additivity"

p3 <- included_full_details |>
  group_by(is_harvest_rate_or_harvest_mortality_estimated_and_reported) |>
  tally() |>
  drop_na() |>
  ggplot(aes(x = is_harvest_rate_or_harvest_mortality_estimated_and_reported, y = n)) +
  geom_histogram(stat = "identity", fill=species_colours[3]) +
  labs(
    x = "Harvest rate estimated",
    y = "Number of records"
  ) +
  theme_minimal() +
  ylim(0, 100) + 
  theme(text = element_text(size = 15))



p4 <- included_full_details |>
  group_by(is_there_an_assessment_of_additivity_of_harvest_mortality) |>
  tally() |>
  drop_na() |>
  ggplot(aes(x = is_there_an_assessment_of_additivity_of_harvest_mortality, y = n)) +
  geom_histogram(stat = "identity", fill=species_colours[2]) +
  labs(
    x = "Assessment of additivity",
    y = ""
  ) +
  theme_minimal() +
  ylim(0, 100) + 
  theme(text = element_text(size = 15))

ggarrange(p3, p4, ncol = 2, labels = c("a)", "b)"), 
          align= "h")


```

From each of the eligible studies, we assessed which outcome variables were examined (@fig-EBV_name2), at two different levels (EBV class and EBV names, respectively). In general, there were about equal number of studies focusing on the EBV classes "Species traits" (67 records) and "Species populations" (60 records). We did not record this variable for simulation based studies. Separating the EBV classes into EBV names, we found that the outcome variable most commonly studied was "Survival" (67 records), followed by "Population density/abundance/growth\*" (47 records). In addition, several studies focused on effects of "Movement rates" or "Habitat use/selection" (12 and 6 records, respectively), and 11 papers reported effects of harvest on "Sex- and age-structure".

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-EBV_name2
#| fig-cap: "The number of studies for each EBV name"


included_full_details |>
  separate_rows(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework, sep = ", ") |>
  group_by(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework) |>
  mutate(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework = case_when(
    outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework %in% c("Behaviour", "Display behaviour/ Gobbling", "Foraging behaviour", "Gobbling activity") ~ "Behaviour",
    TRUE ~ outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework
  )) |>
  mutate(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework = case_when(
    outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework %in% c("Population density *", "Species abundance", "Population growth rate") ~ "Population density/abundance/growth*",
    TRUE ~ outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework
  )) |>
  tally() |>
  drop_na() |>
  mutate(`EBV class` = if_else(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework == "Population density/abundance/growth*", "Species population", "Species trait")) |>
  mutate(`EBV class` = if_else(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework == "Sex- and age structure*", "Species population", `EBV class`)) |>
  ggplot(aes(reorder(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework, n), n)) +
  geom_bar(stat = "identity", aes(fill = `EBV class`)) +
  #geom_histogram(stat = "identity", fill=species_colours[9]) +
  scale_fill_manual(values=c("#5A4A42", "#85361C")) +
  labs(
    x = "EBV name",
    y = "Number of records"
  ) +
  theme_minimal() +
  coord_flip() + 
  theme(legend.position = "top", legend.title = element_blank()) + 
  theme(text = element_text(size = 15))



```

Among studies that were based on (mathematical or statistical) simulations, the most common approach was to compare different harvest strategies (22 records), followed by conducting population viability analyses (17 records) and various forms of sensitivity analyses (13 records). In total, 11 studies were aiming to estimate/calculate maximum sustainable yield. Note that many studies used several of these approaches, so that the combined numbers will be higher than the total of 35 papers that included simulations or scenario analysis.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-simulations
#| fig-cap: "The purpose of simulation studies"

included_full_details |>
  separate_rows(for_simulation_based_scenario_studies_what_is_the_purpose_of_the_simulations, sep = ", ") |> # Split at comma and create new rows
  group_by(for_simulation_based_scenario_studies_what_is_the_purpose_of_the_simulations) |>
  tally() |>
  drop_na() |>
  ggplot(aes(reorder(for_simulation_based_scenario_studies_what_is_the_purpose_of_the_simulations, n), n)) +
  geom_histogram(stat = "identity", fill=species_colours[3]) +
  labs(
    x = "What is the purpose of simulations",
    y = "Number of records"
  ) +
  theme_minimal() +
  coord_flip() + 
  theme(text = element_text(size = 15))

```

The number of studies differed notably among species (@fig-species): The most frequently studied species were wild turkeys (*Meleagris gallopavo*; 33 records), followed by willow ptarmigan/grouse (*Lagopus lagopus*; 28 records), bobwhite (*Colinus virginianus*; 18 records) and sage grouse (*Centrocercus urophasianus*; 18 records). The remaining 11 species that were represented among the eligible papers each had less than 10 papers. Both capercaillie (*Tetrao urogallus*; 4 records) and black grouse (*Lyrurus tetrix*; 6 records) are among these poorly studied species.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-species
#| fig-width: 15
#| fig-cap: "The number of studies for each species and country"


country_data<-included_full_details |>
  separate_rows(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma, sep = ", ") |> # Split at comma and create new rows
  group_by(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma) |>
  mutate(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma = case_when(
    study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma %in% c(
      "Alaska", "California", "Colorado", "Florida", "Georgia","Georiga", "Georgia (USA)", "Idaho", "Illinois", "Indiana", "Kansas", "Kentucky", "Louisiana", "Maine", "Michigan", "Mississippi", "Missouri", "Montana", "Nevada", "New York", "North Carolina", "Ohio",
      "Oklahoma", "Oregon", "Pennsylvania", "Poland", "Rhode Island", "Rhode island", "South Caolina", "South Caraolina", "South Carolina", "South Dakota", "Tennessee", "Texas", "Texas,Texas", "US", "Utah",
      "Virgina", "Virginia", "West Virginia", "Wisconsin", "Wisonsin"
    ) ~ "USA",
    TRUE ~ study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma
  ))|> 
  mutate(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma= case_when(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma %in% c("United Kingdom", "UK", "Scotland", "England")~"UK",
            TRUE~study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma))

country_sp_data<-country_data |> 
  separate_rows(species_if_multiple_seperate_with_a_comma_use_full_latin_names_e_g_lagopus_lagopus_lagopus_muta, sep = ", ") |>
  mutate(
    species_corrected = map_chr(
      species_if_multiple_seperate_with_a_comma_use_full_latin_names_e_g_lagopus_lagopus_lagopus_muta,
      ~ {
        res <- tryCatch(name_backbone(name = .x), error = function(e) NULL)
        if (!is.null(res) && "species" %in% names(res) && !is.na(res$species)) {
          res$species # Return corrected name
        } else {
          .x # Return original if not found
        }
      }
    )
  ) 

# Aggregate data to get counts pr species
species_data <- country_sp_data |> group_by(species_corrected) |>
  tally() |>
  drop_na() |>
  rename(Species = species_corrected)

# Aggregate data to get counts pr species pr country
heatmap_data <- country_sp_data %>%
  count(study_area_location_country_e_g_sweden_if_multiple_seperate_with_a_comma, species_corrected)|> 
  drop_na()

colnames(heatmap_data) <- c("Country", "Species", "Count") 
heatmap_data <- heatmap_data %>% left_join(., species_data)

p7 <- ggplot(species_data, aes(x = reorder(Species, n), y = n, fill = Species)) +
    geom_col() +
    scale_fill_manual(values = species_colours) + 
    labs(
        x = "Species",
        y = "Total number of records"
    ) +
    theme_minimal() +
    coord_flip() + 
    theme(text = element_text(size = 20),
          legend.position = "none")



p8 <- ggplot(heatmap_data, aes(x = Country, y = reorder(Species, n), size = Count, color = Species)) +
    geom_point(alpha = 0.8) +
    scale_color_manual(values = species_colours) + 
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x = "Country", y = "", color = "Species") + 
    theme(legend.position = "none", text = element_text(size = 20))



ggarrange(p7, p8, ncol=2, align = "h")


```

Different species were also subject to research focusing on different outcome variables (@fig-species_ebv). Wild turkey was the most well-studied species in the sense of number of different outcome variables (7 outcome variables), with bobwhite (6 outcome variables) and willow ptarmigan/grouse (5 outcome variables) following closely. Research on capercaillie and black grouse included 3 outcome variables. Two of the outcome variables had only been focused upon in two studies on grey partridge (*Perdix perdix*; effective population size) and rock ptarmigan (*Lagopus muta*; morphology).

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-species_ebv
#| fig-cap: "The distribution of EBV classes by species"

sp_data<-included_full_details |>
  separate_rows(species_if_multiple_seperate_with_a_comma_use_full_latin_names_e_g_lagopus_lagopus_lagopus_muta, sep = ", ") |>
  mutate(
    species_corrected = map_chr(
      species_if_multiple_seperate_with_a_comma_use_full_latin_names_e_g_lagopus_lagopus_lagopus_muta,
      ~ {
        res <- tryCatch(name_backbone(name = .x), error = function(e) NULL)
        if (!is.null(res) && "species" %in% names(res) && !is.na(res$species)) {
          res$species # Return corrected name
        } else {
          .x # Return original if not found
        }
      }
    )
  ) 


sp_EBV<-sp_data |> 
  separate_rows(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework, sep = ", ") |>
  group_by(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework) |>
  mutate(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework = case_when(
    outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework %in% c("Behaviour", "Display behaviour/ Gobbling", "Foraging behaviour", "Gobbling activity") ~ "Behaviour",
    TRUE ~ outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework
  )) 
  
# Aggregate data to get counts
heatmap_data <- sp_EBV |>
  mutate(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework = case_when(
    outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework %in% c("Population density *", "Species abundance", "Population growth rate") ~ "Population density/abundance/growth*",
    TRUE ~ outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework
  )) |>
  count(outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework, species_corrected) |> 
  drop_na()

# Rename columns for easier use in ggplot
colnames(heatmap_data) <- c("EBVname", "Species", "Count")

heatmap_data$Species <- factor(heatmap_data$Species, levels = levels(with(species_data, reorder(Species, n))))
#species_colours
heatmap_data$colour <- species_colours[as.character(heatmap_data$Species)]


ggplot(heatmap_data, aes(x = EBVname, y = Species, size = Count, color = Species)) +
  geom_point(alpha = 0.8) +
 scale_color_manual(values = species_colours)+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for readability
  labs(x = "EBV name",
       y = "Species") +  
  theme(text = element_text(size = 15),
        legend.position = "none")


```

Only 6 studies were based the robust RCT approach, of which 5 were about willow ptarmigan/grouse and one about wild turkeys. Four species (wild turkey, willow and rock ptarmigan and the dusky grouse (*Dendragapus obscurus*)) had been studied using a before-after-control-impact (BACI) approach.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-species_design
#| fig-cap: "The design of observational studies by species"
sp_data<-included_full_details |>
  separate_rows(species_if_multiple_seperate_with_a_comma_use_full_latin_names_e_g_lagopus_lagopus_lagopus_muta, sep = ", ") |>
  mutate(
    species_corrected = map_chr(
      species_if_multiple_seperate_with_a_comma_use_full_latin_names_e_g_lagopus_lagopus_lagopus_muta,
      ~ {
        res <- tryCatch(name_backbone(name = .x), error = function(e) NULL)
        if (!is.null(res) && "species" %in% names(res) && !is.na(res$species)) {
          res$species # Return corrected name
        } else {
          .x # Return original if not found
        }
      }
    )
  ) 


sp_design<-sp_data |> 
  separate_rows(study_design_observational_studies_skip_if_simulation_study_selected_above, sep = ", ") |>
  group_by(study_design_observational_studies_skip_if_simulation_study_selected_above) #|>
  #mutate(study_design_observational_studies_skip_if_simulation_study_selected_above = case_when(
   # outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework %in% c("Behaviour", "Display behaviour/ Gobbling", "Foraging behaviour", "Gobbling activity") ~ "Behaviour",
    #TRUE ~ outcome_variable_ebv_name_not_exhaustive_indicate_name_that_is_not_included_in_original_framework
  #)) 
# Aggregate data to get counts
heatmap_data <- sp_design |> 
  count(study_design_observational_studies_skip_if_simulation_study_selected_above, species_corrected) |>
    mutate(study_design_observational_studies_skip_if_simulation_study_selected_above = fct_recode(study_design_observational_studies_skip_if_simulation_study_selected_above, BACI = "Before-After-Control-Impact (BACI)", RCT = "Randomized Controlled Trials (RCTs)")) |>
mutate(name = fct_relevel(study_design_observational_studies_skip_if_simulation_study_selected_above, 
                            "RCT", "BACI", "Before-After (BA)", 
                            "Control-Impact (CA)", "After (A)")) |>
  drop_na()

# Rename columns for easier use in ggplot
colnames(heatmap_data) <- c("Design", "Species", "Count")
heatmap_data$Species <- factor(heatmap_data$Species, levels = levels(with(species_data, reorder(Species, n))))
#species_colours
heatmap_data$colour <- species_colours[as.character(heatmap_data$Species)]


ggplot(heatmap_data, aes(x = Design, y = Species, size = Count, colour=colour)) +
  geom_point(alpha = 0.8)+
   scale_color_manual(values = unique(heatmap_data$colour))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for readability
  labs(x = "Study design",
       y = "Species",
       fill="") +
  theme(legend.position = "none", legend.title = element_blank()) +
  theme(text = element_text(size = 15))
```

## Summary of the literature mapping

Overall, there has been a substantial amount of research on the effects of harvest on galliform species. However, our mapping of the literature revealed that the literature is dominated by non-experimental research and often conducted at a local or regional scale. We identified only two research projects that were based on an RCT approach, of which one of the projects (represented with five papers) was the former Norwegian project on willow ptarmigan carried out in the late 1990s. Moreover, while the willow ptarmigan (also known as willow grouse) was well represented in the literature (28 records), there was little research on capercaillie (4 records) and black grouse (6 records), and no experimental assessments of the effects of harvest were identified.

The willow ptarmigan study units naturally include various amounts of habitat unsuitable for willow ptarmigan (e.g. lakes, lowland forest in the unit edges, or high mountain peaks). Thus, for willow ptarmigan we calculated area size of suitable habitat within the study units, following the classification by @kvasnes2018quantifying. For forest grouse, the study units are usually smaller and the total area of the selected management units are generally representative of the forest grouse habitat within each unit. Thus, for the forest grouse species we use the total area of the units as basis for the abundance calculations. A complete list of included study units is found in Appendix 3.

# Discussion

The systematic mapping of the literature revealed that, although there has been relatively extensive research on the effects of harvest on galliform species, there are several important key aspects that require further research and emphasis. First, the literature is dominated by research with study designs with a relatively weak ability to detect causal relationships (including After, Before-After and Control-Impact). Although these are often the only options that are available due to ethical, logistic and practical reasons, it is well known that manipulative experiments and in particular Randomized Controlled Trials (RCT) are better suited for drawing firm conclusions about causal relationships. This holds true even though recent advances in statistical methodology also allow to a larger extent to make causal inference from non-manipulative experiments [@christie2019simple]. In our systematic literature mapping, we only identified two research projects that relied on RCT: One study from Norway represented with five publications, and one study from USA on bobwhite. As documenting the relative additivity of harvest mortality under different environmental and demographic conditions is fraught with statistical difficulties, manipulative randomized experiments would be particularly important in order to progress our understanding of this topic.

Second, most studies had a relatively small or moderate spatial extent. As ecological dynamics are often context dependent [@Bowler2020] and might vary both in time and space, and because some ecological properties play out over larger areas, this might limit our ability to draw firm conclusions that could be transferred to other ecological conditions (and species). Our planned study will be unique not only because it has a (near) national spatial extent, but also because it covers three related but ecologically distinct species. While there has been a fair amount of research on willow ptarmigan related to effects of harvest (including the before-mentioned RCTs from the 1990s), the amount of similar research on black grouse and capercaillie is very limited. This highlights that the current management of these species is not based on a solid knowledge base about harvest sustainability, such as for willow ptarmigan management (although we have limited knowledge of harvest sustainability of willow ptarmigan under the current environmental conditions).
